
\documentclass{article}
\usepackage{amsmath}

\oddsidemargin=0in
\evensidemargin=0in
\textwidth=6.3in
\topmargin=-.5in
\textheight=9in

\parindent=0in
\pagestyle{empty}

\input{testpoints}

\begin{document}


% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo
%                             COVER
% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo 

\centerline{\huge \bf Midterm 1}        %%%(number of test)
\vfill \vfill
   
Biostat 701                               %%%(class number and section) 

10/21/2015 \hfill                             %%%(date of test)
{\bf Name: } $\underbrace{\hspace{2.7in}}_{\mbox{\tiny by writing my 
                                           name i swear by the honor code}}$
\vfill \vfill \vfill

{\bf Read all of the following information before starting the exam:}
\vspace{1pc}

\begin{itemize}                        %%%(change info. as desired)
	\item  Show all work, clearly and in order, if you want to get full
	credit.  I reserve the right to take off points if I cannot see how you 
	arrived at your answer (even if your final answer is correct).
	
	\item Justify your answers algebraically whenever possible to ensure 
	full credit. 
		
	\item  This test has 5 problems  %%%(number of problems)
	and is worth 100 points,           %%%(insert total number of points)  
	plus some extra credit at the end. 
	
	\item  Good luck!
\end{itemize}

\vfill \vfill \vfill

\clearpage

% problem with three parts

\begin{problem}{3}
Provide the name and pmf/pdf of the distribution of the following variables:
\begin{enumerate}
\item $X = $ number of independent $Bern(p)$ trials to obtain the first success.
\item $X = $ number of successes in $n$ independent $Bern(p)$ trials.
\item $X = $ number of independent $Bern(p)$ trials to obtain $r$ successes.
\end{enumerate}
\end{problem}
\begin{problem}{5}
Consider the following discrete function:
$$f(n) = \left\{
\begin{matrix}
 \frac{c}{n^3}& \textrm{ for } n=0,1,2,3,...\\
  0 &\textrm{ otherwise }
\end{matrix}\right.$$

Is there a value of $c$ that makes $f$ a valid pmf?
\end{problem}


\begin{problem}{20}
Consider the following (piece-wise) continuous function:
$$f(x) = \left\{
\begin{matrix}
 \frac{c}{x^4}& \textrm{ for } x>1\\
  0 &\textrm{ otherwise }
\end{matrix}\right.$$
\vspace{1pc}

\newpart{10}
Find the value of $c$ that makes this a valid pdf.
\vspace{1pc}

\newpart{10}
Find the CDF of a random variable $X$ with pdf given by $f$.
\end{problem}

\begin{problem}{20}
Let $X \sim \mathcal{N}(\mu,\sigma^2)$.  
\vspace{1pc}

\newpart{10}
Given that the $98^{th}$ percentile of the standard normal distribution is $z_{0.98} = 2.05$, compute the $98^{th}$ percentile of $X$, in terms of $\mu$ and $\sigma$.
\vspace{1pc}

\newpart{5}
Give an expression that represents the probability that $a<X<b$, for $a,b$ real numbers.
\vspace{1pc}

\newpart{5}
What is the probability that $X=\mu$?
\end{problem}
\vspace{1pc}

\begin{problem}{12}
Let $X$ and $Y$ be independent random variables with mean and variance $\mu_X,\sigma_X^2$ and $\mu_Y$ and $\sigma_Y^2$ respectively. Let
$$W= 4X-2Y +3$$
\vspace{1pc}

\newpart{5}
Compute $E(W)$.
\vspace{1pc}

\newpart{7}
Compute $Var(W)$.

\end{problem}

\begin{problem}{20}
Let $X \sim Unif(\left\{0,1,2,...,n\right\})$. Compute the mean and variance of $X$. \\\\

Note that:
$$\sum_{i=1}^n i = \frac{n(n+1)}2$$
$$\sum_{i=1}^n i^2 = \frac{n(n+1)(2n+1)}6$$

\end{problem}


\begin{problem}{20}
My gmail filter divides my email messages into 'Inbox', 'Social' and 'Promotions' (thank you, Google!). I estimate that 50\% of my messages are filtered into my Inbox, 20\% go into Social (no thank you, Facebook) and 30\% go into Promotions (go away, Macy's). I notice that about 90\% of the emails in Promotions contain the word 'free', whereas about 1\% of the messages in Inbox and 2\% of the messages in Social contain the word 'free'. A message comes in with the word 'free'. What is the probability that the message is filtered as Promotion?    
\vspace{1pc}
\end{problem}

% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo
%                         BONUS & SURVEY
% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo

{\large \bf Bonus Question (2 Extra Credit Points):}
\vspace{1pc}

BONUS QUESTION  
\vspace{1pc}
Give the R code for generating $1000$ samples from a normal distribution with mean $1$ and variance $2$, and computing the first sample moment.
\showpoints
\end{document}


