
\documentclass{article}
\usepackage{amsmath}

\oddsidemargin=0in
\evensidemargin=0in
\textwidth=6.3in
\topmargin=-.5in
\textheight=9in

\parindent=0in
\pagestyle{empty}
\newcommand{\pdf}[2]{\left\{\begin{matrix}
{#1} & {#2}\\\\\\0&\textrm{otherwise}
\end{matrix}\right.}
\newcommand{\infi}{\int_{-\infty}^\infty}

\input{testpoints}

\begin{document}


% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo
%                             COVER
% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo 

\centerline{\huge \bf Midterm 1}        %%%(number of test)
\vfill \vfill
   
Biostat 701                               %%%(class number and section) 

11/18/2015 \hfill                             %%%(date of test)
{\bf Name: } $\underbrace{\hspace{2.7in}}_{\mbox{\tiny by writing my 
                                           name i swear by the honor code}}$
\vfill \vfill \vfill

{\bf Read all of the following information before starting the exam:}
\vspace{1pc}

\begin{itemize}                        %%%(change info. as desired)
	\item  Show all work, clearly and in order, if you want to get full
	credit.  I reserve the right to take off points if I cannot see how you 
	arrived at your answer (even if your final answer is correct).
	
	\item Justify your answers algebraically whenever possible to ensure 
	full credit. 
		
	\item  This test has 5 problems  %%%(number of problems)
	and is worth 100 points,           %%%(insert total number of points)  
	plus some extra credit at the end. 
	
	\item  Good luck!
\end{itemize}

\vfill \vfill \vfill

\clearpage

% problem with three parts

\begin{problem}{20}
Let $\sim Exp(\lambda)$.
\vspace{1pc}

\newpart{10}

Compute the moment generating function for $X$.
\vspace{1pc}

\newpart{10}

Use the MGF in part a to compute the mean and variance of $X$.
\vspace{1pc}

\end{problem}
\begin{problem}{15}
Let $X_1,X_2$ be continuous random variables with joint pdf 
$$f(x_1,x_2) = \pdf{4x_1x_2}{0<x_1<1 \textrm{ and } 0<x_2<1}$$

\vspace{1pc}

\newpart{10}
Compute the marginal pdfs of $X_1$ and $X_2$.

\vspace{1pc}

\newpart{5}
Are $X_1$ and $X_2$ independent? Why or why not?

\end{problem}


\begin{problem}{25}
Let $X,Y \sim BVN$.

\vspace{1pc}

\newpart{15}
Define all the relevant parameters and write the joint pdf of $X$ and $Y$

\vspace{1pc}

\newpart{5}

Let $Z\sim N(0,1)$. True or False: $X,Y,Z \sim MVN$.

\vspace{1pc}

\newpart{5}

Let $W\sim N(\mu_w,\sigma^2_w)$ and suppose that $corr(X,W) = corr(Y,W) = 0$. What is the pdf of $X,Y,W$?
 
\end{problem}

\begin{problem}{20}
Let $X_1,X_2$ be i.i.d $Exp(1)$. Consider the transformation:
$$Y_1 = X_1 - X_2$$
$$Y_2 = X_1 + X_2$$

Find $f_{Y_1Y_2}(y_1,y_2)$. Don't forget to specify the support! (Consider the support of $x_1,x_2$ and transform to make a statement about the bounds on $y_1,y_2$. 
\end{problem}

\vspace{1pc}

\begin{problem}{20}
Consider the 'seed example' from class: A triangular area of land is under observation. The probability that an airborne seed lands anywhere in the region is modeled by a uniform distribution on the region. I.e.,$0<x<2$ and $0<y<x/2$. In class, we computed 
$$f_{Y|x}(y) = \pdf{\frac2{x}}{0<y<\frac{x}2}$$  
and the marginals are:\\
$$f_X(x) = \pdf{\frac{x}2}{0<x<2}$$
$$f_Y(y) = \pdf{2(1-y)}{0<y<1}$$
\vspace{1pc}

\newpart{10}
What is the random variable $E(Y|X)$?
\vspace{1pc}

\newpart{10}
Compute $E(E(Y|X))$.

\end{problem}


% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo
%                         BONUS & SURVEY
% ooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooooo

\showpoints
\end{document}


